Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2015-01-24 algorithm.tex
AA: dato un colpo di aspell. Manca rilettura finale. Da rivedere la caption della figura 10

\section{PS algorithm}
\secL{algorithm}
In this section we present PS, the algorithm that we propose to perform the Design Space Exploration (DSE). As any other DSE algorithm, its input is the parameterized system $S$ and a benchmark application $b$, while the output is an approximation of the Pareto-optimal set $\mathcal{P}\left(S,b\right)$.   
PS is iterative and we will describe, separately, the initialization phase, the operations performed inside each iteration and the condition that triggers the termination.

@@

With regard to the performance of the proposed approach, we observe that the
performance of $PS$ is related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evaluating a new
configuration is the time-dominating one among all the operations
performed. As we will show in the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark
compilation and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  The $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, the number of total new simulations $N_{tot}$ 
is bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the number of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
pareto $\mathscr{P}_0$.

>>>>

\section{PS algorithm}
\secL{algorithm}
In this section we present PS, the algorithm that we propose to perform the Design Space Exploration (DSE). As any other DSE algorithm, its input is the parametrized system $S$ and a benchmark application $b$, while the output is an approximation of the Pareto-optimal set $\mathcal{P}\left(S,b\right)$.   
PS is iterative and we will describe, separately, the initialization phase, the operations performed inside each iteration and the condition that triggers the termination.

@@

With regard to the performance of the proposed approach, we observe that the
performance of $PS$ is related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evaluating a new
configuration is the time-dominating one among all the operations
performed. As we will show in the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark
compilation and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  The $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, the number of total new simulations $N_{tot}$ 
is bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the number of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
Pareto $\mathscr{P}_0$.

### 2015-01-24 concepts.tex
AA: dato un colpo di aspell. Manca rilettura finale. Da rivedere la caption della figura 10

Let $S$ be a parameterized system with $n$ parameters. The generic
parameter $p_i, \ i \in \{1,2,\ldots,n\}$ can take any value in
the set $V_i$. A {\em configuration} $\mathbf{c}$ of the system
$S$ is a $n$-tuple $\langle v_1,v_2,\ldots,v_n \rangle$ in which
$v_i \in V_i$ is the value fixed for the parameter $p_i$. The {\em
configuration space} (or {\em design space}) of $S$ [which we will
indicate as $\mathcal{C}(S)$] is the complete range of possible
configurations [$\mathcal{C}(S) = V_1 \times V_2 \times \ldots
\times V_n$]. Naturally, not all the configurations of
$\mathcal{C}(S)$ can really be mapped on $S$, but only a subset 
is feasible. We call this subset {\em
feasible configuration space} of $S$ and indicate it as
$\mathcal{C}^*(S)$.

@@

In the design process, there are often a lot of ``local decisions'' in
which the designer can take advantage of well-known results or
heuristics. For example, if we consider a highly parallel
computer architecture in a configuration with very few registers, it is
well-known that increasing the number of ALUs does not bring tangible
better performances but only cause an increase in area occupation. An
intelligent exploration algorithm would not waste time evaluating
configurations with many ALUs and few registers. If $p_{1}$ is the
parameter ``number of registers'' and $p_{2}$ is the parameter
``number of ALUs'', we can say that the region:
\[
R=\left\{ \left.\left(v_{1},v_{2}\right)\in V_{1}\times V_{2}\right|v_{1}<s_{1},v_{2}>s_{2}\right\} 
\]
 (where $s_{1}$ and $s_{2}$ are some threshold values) is \emph{uninteresting}. This clearly shows that there are cases when, if a parameter lies in
a certain range, there are ranges of the other one that are not worth being explored. Intuitively, the cartesian product of these ranges is an \emph{uninteresting region}.

@@

Since regions will be defined as cartesian products of parameter intervals, we need the following definition.
\begin{definition}[Parameter interval] Let $p_{i}$ be a parameter and $a_{i},b_{i}\in V_{i},a_{i}<b_{i}$.
A $p_{i}$- interval is
	\[
	\left[a_{i}..b_{i}\right]=\left[a_{i},b_{i}\right]\cap V_{i}
	\]
i.e. taking the interval $\left[a_{i}..b_{i}\right]$
is the set of values of $V_{i}$ lying in $\left[a_{i},b_{i}\right]$)
\end{definition}

@@

We consider only parameters $p_{i}$ with ordering, i.e. such
that $\forall a,b\in V_{i}$ it is possible
to say $a<b$ or $a=b$ or $a>b$. 
The following definition of interval contiguity will be the base of the concept of region contiguity, used to define merging and splitting operations.
\begin{definition}[Contiguous intervals]
\label{pers02.def:Contiguous-intervals}
Two intervals are contiguous iff they do not overlap and merging them results in a new interval. More formally, let $\left[a_{i}..b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$
be two $p_{i}$-intervals. They are said to be contiguous (see \figR{interval_and_region}(a) ) iff
\begin{align}\begin{array}{l}
	\left[a_{i}..b_{i}\right]\cap\left[x_{i}..y_{i}\right]=\emptyset \mbox{      and} \\
	\left[a_{i}..b_{i}\right]\cup\left[x_{i}..y_{i}\right]$ is a $p_{i}$-
interval or $\left[x_{i}..y_{i}\right]\cup\left[a_{i}..b_{i}\right]
\end{array}\end{align}
\end{definition}

\begin{definition}[Region] 
A region is a cartesian product of parameter intervals, as illustrated in \figR{interval_and_region} (b). More formally, if $p_1,\dots,p_n$ are the parameters of the system, and $\left[ a_{i}..b_{i} \right]$ is a $p_i$-interval, a region is a set of the following form:
\[
R=\left[a_{1}..b_{1}\right]\times\dots\times\left[a_{n}..b_{n}\right]=\prod\left[a_{i}..b_{i}\right]
\]
\end{definition}

We will refer later to ``big'' and ``small'' regions. The sense
of these terms is related to the cardinality of regions.
\begin{definition}[Interval and region comparison]
A $p_{i}$-interval $\left[a_{i}..b_{i}\right]$
is said to be bigger then another $p_{i}$-interval $\left[c_{i}..d_{i}\right]$
iff 
$\left|\left[a_{i}..b_{i}\right]\right|>\left|\left[c_{i}..d_{i}\right]\right|$.
Similarly, a region $R_{1}$ is said to be \emph{bigger }then $R_{2}$, if
$\left|R_{1}\right|>\left|R_{2}\right|$
\end{definition}
Note that with $|\cdot|$ we indicate the cardinality of a set. Therefore, $\left|\left[a_{i}..b_{i}\right]\right|$ is the number of values of $V_i$ lying between $a_{i}$ and $b_{i}$ and has not to be confused with $b_i - a_i$.
Similarly $\left|R_{i}\right|$ is the number of configurations inside the region $R_i$.

@@

\begin{definition}[Splitting a region]
\defL{splitting-regions}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it into a set of sub-regions obtained splitting each $p_i$-interval and combining the resulting sub-intervals with the cartesian product, as shown in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

@@

The merging operation will be defined only on region pairs that satisfy the conditions provided in the following definition.
\begin{definition}[Contiguous regions]Let $R_{1}=\prod\left[a_{i}\dots b_{i}\right]$ and $R_{2}=\prod\left[x_{i}..y_{i}\right]$
be two regions. They are said to be contiguous iff they do not overlap and, merging them, a new region can be obtained. Formally, $R_1$ and $R_2$ are contiguous iff there exists a $j$ such that:
	\begin{align}\begin{array}{l}
		\left[a_{i}..b_{i}\right]=\left[x_{i}..y_{i}\right]$
	\mbox{ for } $i\neq j \mbox{      and} \\
		\left[a_{j}..b_{j}\right]$ and $\left[x_{j}..y_{j}\right] \mbox{ are contiguous intervals}
	\end{array}\end{align}
\noindent We call $p_j$ the \emph{contiguity parameter} for the two regions.
\end{definition}

Examples of contiguous and non contiguous regions are given in \figR{contiguous_regions}. The two regions at the top are contiguous because their union can be expressed as the cartesian product of parameter intervals, thus they can be merged to obtain a new bigger region. The same also applies to the regions on the right. On the contrary, the two regions on the left (or of the two regions at the bottom) cannot be merged.

>>>>

Let $S$ be a parametrized system with $n$ parameters. The generic
parameter $p_i, \ i \in \{1,2,\ldots,n\}$ can take any value in
the set $V_i$. A {\em configuration} $\mathbf{c}$ of the system
$S$ is a $n$-tuple $\langle v_1,v_2,\ldots,v_n \rangle$ in which
$v_i \in V_i$ is the value fixed for the parameter $p_i$. The {\em
configuration space} (or {\em design space}) of $S$ [which we will
indicate as $\mathcal{C}(S)$] is the complete range of possible
configurations [$\mathcal{C}(S) = V_1 \times V_2 \times \ldots
\times V_n$]. Naturally, not all the configurations of
$\mathcal{C}(S)$ can really be mapped on $S$, but only a subset 
is feasible. We call this subset {\em
feasible configuration space} of $S$ and indicate it as
$\mathcal{C}^*(S)$.

@@

In the design process, there are often a lot of ``local decisions'' in
which the designer can take advantage of well-known results or
heuristics. For example, if we consider a highly parallel
computer architecture in a configuration with very few registers, it is
well-known that increasing the number of ALUs does not bring tangible
better performances but only cause an increase in area occupation. An
intelligent exploration algorithm would not waste time evaluating
configurations with many ALUs and few registers. If $p_{1}$ is the
parameter ``number of registers'' and $p_{2}$ is the parameter
``number of ALUs'', we can say that the region:
\[
R=\left\{ \left.\left(v_{1},v_{2}\right)\in V_{1}\times V_{2}\right|v_{1}<s_{1},v_{2}>s_{2}\right\} 
\]
 (where $s_{1}$ and $s_{2}$ are some threshold values) is \emph{uninteresting}. This clearly shows that there are cases when, if a parameter lies in
a certain range, there are ranges of the other one that are not worth being explored. Intuitively, the Cartesian product of these ranges is an \emph{uninteresting region}.

@@

Since regions will be defined as Cartesian products of parameter intervals, we need the following definition.
\begin{definition}[Parameter interval] Let $p_{i}$ be a parameter and $a_{i},b_{i}\in V_{i},a_{i}<b_{i}$.
A $p_{i}$- interval is
	\[
	\left[a_{i}..b_{i}\right]=\left[a_{i},b_{i}\right]\cap V_{i}
	\]
i.e. taking the interval $\left[a_{i}..b_{i}\right]$
is the set of values of $V_{i}$ lying in $\left[a_{i},b_{i}\right]$)
\end{definition}

@@

We consider only parameters $p_{i}$ with ordering, i.e. such
that $\forall a,b\in V_{i}$ it is possible
to say $a<b$ or $a=b$ or $a>b$. 
The following definition of interval contiguity will be the base of the concept of region contiguity, used to define merging and splitting operations.
\begin{definition}[Contiguous intervals]
\label{pers02.def:Contiguous-intervals}
Two intervals are contiguous if and only if they do not overlap and merging them results in a new interval. More formally, let $\left[a_{i}..b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$
be two $p_{i}$-intervals. They are said to be contiguous (see \figR{interval_and_region}(a) ) if and only if
\begin{align}\begin{array}{l}
	\left[a_{i}..b_{i}\right]\cap\left[x_{i}..y_{i}\right]=\emptyset \mbox{      and} \\
	\left[a_{i}..b_{i}\right]\cup\left[x_{i}..y_{i}\right]$ is a $p_{i}$-
interval or $\left[x_{i}..y_{i}\right]\cup\left[a_{i}..b_{i}\right]
\end{array}\end{align}
\end{definition}

\begin{definition}[Region] 
A region is a Cartesian product of parameter intervals, as illustrated in \figR{interval_and_region} (b). More formally, if $p_1,\dots,p_n$ are the parameters of the system, and $\left[ a_{i}..b_{i} \right]$ is a $p_i$-interval, a region is a set of the following form:
\[
R=\left[a_{1}..b_{1}\right]\times\dots\times\left[a_{n}..b_{n}\right]=\prod\left[a_{i}..b_{i}\right]
\]
\end{definition}

We will refer later to ``big'' and ``small'' regions. The sense
of these terms is related to the cardinality of regions.
\begin{definition}[Interval and region comparison]
A $p_{i}$-interval $\left[a_{i}..b_{i}\right]$
is said to be bigger then another $p_{i}$-interval $\left[c_{i}..d_{i}\right]$
if and only if 
$\left|\left[a_{i}..b_{i}\right]\right|>\left|\left[c_{i}..d_{i}\right]\right|$.
Similarly, a region $R_{1}$ is said to be \emph{bigger }then $R_{2}$, if
$\left|R_{1}\right|>\left|R_{2}\right|$
\end{definition}
Note that with $|\cdot|$ we indicate the cardinality of a set. Therefore, $\left|\left[a_{i}..b_{i}\right]\right|$ is the number of values of $V_i$ lying between $a_{i}$ and $b_{i}$ and has not to be confused with $b_i - a_i$.
Similarly $\left|R_{i}\right|$ is the number of configurations inside the region $R_i$.

@@

\begin{definition}[Splitting a region]
\defL{splitting-regions}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it into a set of sub-regions obtained splitting each $p_i$-interval and combining the resulting sub-intervals with the Cartesian product, as shown in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

@@

The merging operation will be defined only on region pairs that satisfy the conditions provided in the following definition.
\begin{definition}[Contiguous regions]Let $R_{1}=\prod\left[a_{i}\dots b_{i}\right]$ and $R_{2}=\prod\left[x_{i}..y_{i}\right]$
be two regions. They are said to be contiguous if and only if they do not overlap and, merging them, a new region can be obtained. Formally, $R_1$ and $R_2$ are contiguous if and only if there exists a $j$ such that:
	\begin{align}\begin{array}{l}
		\left[a_{i}..b_{i}\right]=\left[x_{i}..y_{i}\right]$
	\mbox{ for } $i\neq j \mbox{      and} \\
		\left[a_{j}..b_{j}\right]$ and $\left[x_{j}..y_{j}\right] \mbox{ are contiguous intervals}
	\end{array}\end{align}
\noindent We call $p_j$ the \emph{contiguity parameter} for the two regions.
\end{definition}

Examples of contiguous and non contiguous regions are given in \figR{contiguous_regions}. The two regions at the top are contiguous because their union can be expressed as the Cartesian product of parameter intervals, thus they can be merged to obtain a new bigger region. The same also applies to the regions on the right. On the contrary, the two regions on the left (or of the two regions at the bottom) cannot be merged.

### 2015-01-24 experiments.tex
AA: errori segnalati da reviewer 2 corretti. Rilettura ancora da fare
AA: dato un colpo di aspell. Manca rilettura finale. Da rivedere la caption della figura 10

In particular, we will use a parameterized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Software Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

In this section, we will also briefly describe the parameterized platform used
as testbed for the experiments, the general evaluation flow along with
the high-level estimation models and applications used as benchmarks.

@@

Although an analogy between the concepts of ``era'' and ``generation''
can be considered, comparisons have been carried out using as
reference a fixed amount of total unique simulations, instead of a given
number of eras/generations . The purpose is to make the comparison
fair and accurate from the algorithm performance perspective, since a fixed budget of
different simulations is more directly related to the actual amount of time
required to execute the exploration.
%; on the other side, simply
%referring to a given number of generations/eras could be inaccurate
%when a considerable number of configurations remain unaltered in the
%current population (e.g. because of a local optimum)
\comment{AA: Da 'on the other side' fin qui non ho capito cosa vuoi dire}.
\comment{DAV: se non lo capisci vuol dire che o e' scritto male o e'
inutile, per adesso lo tolgo.}

@@

\begin{figure}
  \figLC{kimpact}{Impact of increased K values on Pareto fronts and distribution for the alloca\_test application. \comment{AA: non capisco cosa sia la figura a destra. Inoltre, io vedo sulle ascisse e le ordinate di entrambi i grafici, le stesse grandezze (Time vs Power), quindi mi aspetto di trovare esattamente lo stesso plot sia a destra che a sinistra}}
  \begin{center}
    \includegraphics[width=0.45\textwidth]{pictures/alloca_K.eps} 
    \includegraphics[width=0.45\textwidth]{pictures/alloca_K_ALL.eps} 
  \end{center}
\end{figure}

@@

\begin{figure}
  \figLC{dispersion}{(a)(b) Variation range obtained by PS and GA for
  different benchmarks and average dispersion absolute dispersion (c).
  \comment{AA: manca il titolo nella figura (c)} }
  \begin{center}
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} }
  \end{center}
\end{figure}

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and (b)
confirm the of density vs extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secL{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentiality of the designed device.

>>>>

In particular, we will use a parametrized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Software Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

In this section, we will also briefly describe the parametrized platform used
as testbed for the experiments, the general evaluation flow along with
the high-level estimation models and applications used as benchmarks.

@@

Although an analogy between the concepts of ``era'' and ``generation''
can be considered, comparisons have been carried out using as
reference a fixed amount of total unique simulations, instead of a given
number of eras/generations . The purpose is to make the comparison
fair and accurate from the algorithm performance perspective, since a fixed budget of
different simulations is more directly related to the actual amount of time
required to execute the exploration.
%; on the other side, simply
%referring to a given number of generations/eras could be inaccurate
%when a considerable number of configurations remain unaltered in the
%current population (e.g. because of a local optimum)

@@

\begin{figure}
  \figLC{kimpact}{Impact of increased K values on Pareto fronts and distribution for the alloca\_test application. }
  \begin{center}
    \includegraphics[width=0.45\textwidth]{pictures/alloca_K.eps} 
    \includegraphics[width=0.45\textwidth]{pictures/alloca_K_ALL.eps} 
  \end{center}
\end{figure}

@@

\begin{figure}
  \figLC{dispersion}{(a)(b) Variation range obtained by PS and GA for
  different benchmarks and average dispersion absolute dispersion (c). }
  \begin{center}
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} }
    \subfigure[]{\includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} }
  \end{center}
\end{figure}

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and (b)
confirm the of density vs extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secR{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentiality of the designed device.
